Distributional Trees and Forests for Circular Data

### 2019-07-03 pkg/disttree/inst/IWSM2019/slides.Rnw
revision of slides for IWSM 2019, added map and boxplots
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2638 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 8a30f663555be7c4a2f53bc2c8cb2a058d97fee7

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
% \usepackage{Sweave}
\usepackage{amsmath,tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
\usepackage{xcolor}
\usepackage{BOONDOX-cal}
%\graphicspath{{plots/}}

@@

# setwd("~/svn/partykit/pkg/disttree/inst/IWSM2019")
library("disttree")
#library("Formula")
library("latex2exp")

@

@@

\begin{frame}[fragile]
\frametitle{Motivation}
\textbf{Goal:} Regression model for a circular response variable%, \\
%\hspace{1.05cm} i.e. a periodic response variable measured on the unit circle.

@@

The density for $y \in [0, 2\pi)$ and parameter vector $\theta = (\mu, \kappa)$ is given by
$$
f_\mathcal{vM}(y; \theta) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
$$
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$.

The corresponding log-likelihood function is defined by
$$
\ell(\theta; y) =  \log(f_\mathcal{vM}(y;\theta)).
$$
\end{frame}

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.7\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit global distributional model $\mathcal{vM}(Y; \theta)$: \\ % to the whole data set:\\
Estimate $\hat{\theta}$ via maximum likelihood\\
$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.23\textwidth}
%\vspace{-0.1cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
}
\visible<4>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\mathcal{vM}(Y;\hat{\theta}$)};
}
\visible<5->{
\node[inner sep=0pt] (density_root) at (0.8, 1.7)
    {\includegraphics[width=0.6\textwidth]{density_root.pdf}};
}
\visible<6-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1);
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2);
}
\visible<9-10>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<11>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{vM}(Y_1;\hat{\theta_1}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.7, 0) {$\mathcal{vM}(Y_2;\hat{\theta_2}$)};
}
\visible<11->{
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<12->{
\node[inner sep=0pt] (density2) at (-0.1,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_left.pdf}};
}
\visible<12->{
\node[inner sep=0pt] (density3) at (1.7,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_right.pdf}};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
%\begin{adjustwidth}{-0.0em}{-1em}
\begin{enumerate}
\setcounter{enumi}{2}
\item<8-> Split along the covariate $X$ with strongest association or instability and\\ at breakpoint $p$ with highest improvement in log-likelihood.
\item<10-> Repeat steps 1--3 recursively until some stopping criterion is met,\\
yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}
%\end{adjustwidth}
\vspace{0.3cm}
\visible<13->{
{\bf Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
}
\end{frame}

@@

%\textbf{Goal:} Predict wind directions for the next hour at Innsbruck airport (Austria).

\textbf{Input:} Observation data of various meteorological quantities of the previous hour(s) from Innsbruck airport and 5 nearby weather stations.

@@

\textbf{Base variables:}
\begin{itemize}
\item Wind direction
\item Wind speed
\item Temperature
\item Air pressure
\item \dots 
\item \textit{FIXME: list of all included covariates}
\end{itemize}
%\begin{center}
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{}
%\end{center}
\end{frame}

%\begin{frame}[fragile]
%\includegraphics{airport_ibk.jpg}
%\end{frame}

@@

\textbf{Distributional tree model:}

\begin{center}
%\vspace*{1cm}
\setkeys{Gin}{width=\linewidth}
\includegraphics{schlosser-circtree_plot.pdf}
\end{center}

@@

\textbf{Data:} .. observations of response variables and .. covariates

\textbf{Cross validation:} 
\begin{itemize}
\item 5-fold setting 
\item learning data: .. observations
\item testing data: .. observations
\end{itemize}

\textbf{Evaluation:} Results are compared based on circular CRPS.

@@

boxplot CRPS cross validation
%\begin{center}
%\vspace*{-1cm}
%\setkeys{Gin}{width=0.8\linewidth}
%\includegraphics{}
%\end{center}
\end{frame}

@@

Breiman L (2001).
 \dquote{Random Forests.}
 \textit{Machine Learning}, 
 \textbf{45}(1), 5--32.
 \doi{10.1023/A:1010933404324}

@@

@@

Mulder K, Klugkist I (2017).
 \dquote{Bayesian Estimation and Hypothesis Tests for a Circular Generalized Linear Model.}
 \textit{Journal of Mathematical Psychology}, 
 \textbf{80}, 4--14. 
 \doi{10.1016/j.jmp.2017.07.001}

>>>>

\setbeamertemplate{caption}{\insertcaption} 
%% includes a replacement for \usepackage{Sweave}
% \usepackage{Sweave}
\usepackage{amsmath,tikz}
\usetikzlibrary{positioning,shapes,arrows,decorations.pathreplacing,calc,automata,mindmap}
\usepackage{xcolor}
%\usepackage[cal=boondoxo]{mathalfa}
%\graphicspath{{plots/}}

@@

# setwd("~/svn/partykit/pkg/disttree/inst/IWSM2019")
library("disttree")
#library("Formula")
library("latex2exp")
library("circmax")
@

@@

\begin{frame}[fragile]
\frametitle{Motivation}
\textbf{Goal:} Regression model for a circular response variable.%, \\
%\hspace{1.05cm} i.e. a periodic response variable measured on the unit circle.

@@

The density for $y \in [0, 2\pi)$ and parameter vector $\theta = (\mu, \kappa)$ is given by
$$
f_\text{vM}(y; \theta) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
$$
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$.

The corresponding log-likelihood function is defined by
$$
\ell(\theta; y) =  \log(f_\text{vM}(y;\theta)).
$$
\end{frame}

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.7\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit global distributional model $\text{vM}(Y; \theta)$: \\ % to the whole data set:\\
Estimate $\hat{\theta}$ via maximum likelihood\\
$\hat{\theta} = \argmax_{\theta \in \Theta} \sum_{i=1}^n \ell(\theta; y_i)$
\item<7-> Test for associations/instabilities of the scores $\frac{\partial \ell}{\partial \theta}(\hat{\theta};y_i)$ and each covariate $X_i$.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.23\textwidth}
%\vspace{-0.1cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$Y$};
}
\visible<4>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=71pt, minimum height = 30pt] (n0) at (0.8, 1.7) {$\text{vM}(Y;\hat{\theta}$)};
}
\visible<5->{
\node[inner sep=0pt] (density_root) at (0.8, 1.7)
    {\includegraphics[width=0.6\textwidth]{density_root.pdf}};
}
\visible<6-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {?};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1);
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2);
}
\visible<9-10>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$Y_2$};
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<11>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n1) at (0, 0) {$\text{vM}(Y_1;\hat{\theta}_1$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 20pt] (n2) at (1.7, 0) {$\text{vM}(Y_2;\hat{\theta}_2$)};
}
\visible<11->{
\draw[-, gray, line width=0.5pt] (0.7, 1.1) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, gray, line width=0.5pt] (1.0, 1.1) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<12->{
\node[inner sep=0pt] (density2) at (-0.1,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_left.pdf}};
}
\visible<12->{
\node[inner sep=0pt] (density3) at (1.7,-0.3)
    {\includegraphics[width=0.5\textwidth]{density_right.pdf}};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
%\begin{adjustwidth}{-0.0em}{-1em}
\begin{enumerate}
\setcounter{enumi}{2}
\item<8-> Split along the covariate $X$ with strongest association or instability and\\ at breakpoint $p$ with highest improvement in log-likelihood.
\item<10-> Repeat steps 1--3 recursively until some stopping criterion is met,\\
yielding $B$ subgroups $\mathcal{B}_b$ with $b = 1, \dots, B$.
\end{enumerate}
%\end{adjustwidth}
\vspace{0.3cm}
\visible<13->{
{\bf Forest:} Ensemble of $T$ trees.
\begin{itemize}
\item Bootstrap or subsamples.
\item Random input variable sampling.
\end{itemize}
}
\end{frame}

@@

\textbf{Response:} Wind direction in the next hour at Innsbruck airport (Austria).

\only<2>{
%\vspace{-0.4cm}
\begin{center}
\setkeys{Gin}{width=0.65\linewidth}
\includegraphics{ibk_airport.png}
\end{center}
}

\only<3->{
\medskip

\textbf{Input:} Observation data of various meteorological quantities from Innsbruck airport and 5 nearby weather stations.

@@

\textbf{Base variables:}
\begin{itemize}
\item Wind direction
\item Wind speed
\item Temperature
\item Air pressure
\item Humidity
% "dd"   "ff"   "ffx"  "p"    "pred" "rf"   "tl"  
\end{itemize}

\textbf{Variations:}
Minima, maxima, changes over time or over location, \ldots

\quad $\rightarrow$ 354 covariates
}
\end{frame}

@@

\textbf{Circular distributional tree:}

\vspace{-1cm}

\setkeys{Gin}{width=1.0\linewidth}

\begin{center}
\includegraphics{schlosser-circtree_plot.pdf}
\end{center}

@@

\textbf{Evaluation setting:} 
\begin{itemize}
\item Data set of $40\ 000$ observations with 354 covariates
\item 5-fold cross validation
\item Results are compared based on circular CRPS (Grimit et al., 2006)
\end{itemize}

@@

\begin{center}
%\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
\includegraphics{boxplot_crpsraw.pdf}
\end{center}
\end{frame}

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Wind Directions}
\begin{center}
%\vspace*{-1cm}
\setkeys{Gin}{width=0.8\linewidth}
\includegraphics{boxplot_crpsskill.pdf}
\end{center}
\end{frame}

@@

%Breiman L (2001).
% \dquote{Random Forests.}
% \textit{Machine Learning}, 
% \textbf{45}(1), 5--32.
% \doi{10.1023/A:1010933404324}
% 
%\medskip

@@

Grimit E P, Gneiting T, Berrocal V J, Johnson N A (2006).
  \dquote{The Continuous Ranked Probability Score for Circular Variables and its Application to Mesoscale Forecast Ensemble Verification.}
  \textit{Q.J.R. Meteorol. Soc.}, 
  \textbf{132}, 2925--2942. 
  \doi{10.1256/qj.05.235}

@@

%Mulder K, Klugkist I (2017).
% \dquote{Bayesian Estimation and Hypothesis Tests for a Circular Generalized Linear Model.}
% \textit{Journal of Mathematical Psychology}, 
% \textbf{80}, 4--14. 
% \doi{10.1016/j.jmp.2017.07.001}
%
%\medskip
